---
title: "Regular expressions in R - solutions to in-class practice"
date: "`r lubridate::today()`"
output: pdf_document
---

```{r}
library(stringr)
suppressPackageStartupMessages(library(tidyverse))
```


**PRACTICE 1.** Use dots and `$` to match the last four characters of all sentences in `x`: 

```{r}
x <- c("This is a sentence to practice regular expressions", 
       "Find the word regex in this sentence",
       "A last sentence similar to the first sentence")
```

```{r practice1-solution}
str_view(x, "....$")
```


**PRACTICE 2.** This sentence contains a set of ugly characters. Identify them first using the `|` operator, then using the `[]`:

```{r}
j <- "This&is#MACS%30500"
```

```{r practice2-solution}
str_view_all(j, "&|#|%")
str_view_all(j, "[&#%]")
```


**PRACTICE 3.** Apply these meta-characters to extract all phone numbers from the following string:

```{r}
s <- "My number is 444-3340, Carl number is 555-3755, Richard number is 666-6681"
```

```{r pratice3-solution}
str_view_all(s, "[0-9][0-9][0-9]-[0-9][0-9][0-9][0-9]")
str_view_all(s, "\\d\\d\\d-\\d\\d\\d\\d")
str_view_all(s, r"(\d\d\d-\d\d\d\d)")
```


**PRACTICE 4.** Match all three emails in the following string:

```{r}
email <- "smith90@uchicago.edu nardin@uchicago.edu happy2@uchicago.edu"
```

```{r pratice4-solution}
str_view_all(email, "[a-z]+[0-9]*@uchicago\\.edu")
```

Solution: `[a-z]+` matches one or more characters, `[0-9]*` matches zero or more numbers, `@uchicago\.edu` is a literal match with `\` to match the dot literally


**PRACTICE 5.** Apply regular expressions to "Alice in Wonderland"

```{r}
alice <- "Alice was beginning to get very tired of sitting by her sister on the bank, and of having nothing to do: \
once or twice she had peeped into the book her sister was reading, but it had no pictures or conversations in it, \
'and what is the use of a book,' thought Alice 'without pictures or conversation?' So she was considering in her own mind \
(as well as she could, for the hot day made her feel very sleepy and stupid), whether the pleasure of making a daisy-chain \
would be worth the trouble of getting up and picking the daisies, when suddenly a White Rabbit with pink eyes ran close by her.\
There was nothing so VERY remarkable in that; nor did Alice think it so VERY much out of the way to hear the Rabbit say to \
itself, 'Oh dear!  Oh dear!  I shall be late!' (when she thought it over afterwards, it occurred to her that she ought to have \
wondered at this, but at the time it all seemed quite natural); but when the Rabbit actually \
TOOK A WATCH OUT OF ITS WAISTCOAT-POCKET, and looked at it, and then hurried on, Alice started to her feet, \
for it flashed across her mind that she had never before seen a rabbit with either a waistcoat-pocket, or a watch to \
take out of it, and burning with curiosity, she ran across the field after it, and fortunately was just in time to see it pop \
down a large rabbit-hole under the hedge. \
In another moment down went Alice after it, never once considering how in the world she was to get out again. \
The rabbit-hole went straight on like a tunnel for some way, and then dipped suddenly down, so suddenly that \
Alice had not a moment to think about stopping herself before she found herself falling down a very deep well. \
Either the well was very deep, or she fell very slowly, for she had plenty of time as she went down to look about her and to \
wonder what was going to happen next.  First, she tried to look down and make out what she was coming to, \
but it was too dark to see anything; then she looked at the sides of the well, and noticed that they were filled with \
cupboards and book-shelves; here and there she saw maps and pictures hung upon pegs. \
She took down a jar from one of the shelves as she passed; it was labelled `ORANGE MARMALADE', \
but to her great disappointment it was empty:  she did not like to drop the jar for fear of killing somebody, \
so managed to put it into one of the cupboards as she fell past it. \ "
```

Find and replace all uppercase words. Notice `\\b` specifies word boundaries (word boundary anchor) 
```{r}
# find
str_view(alice, "\\b[A-Z]+\\b")

# replace
str_replace_all(alice, "\\b[A-Z]+\\b", function(x) tolower(x))
```

Find all present progressive verbs that have as previous word an auxiliar (e.g., was beginning, was reading)
```{r}
pattern <- "\\b(is|are|was|were)\\s+[a-z]+ing\\b"
matches <- str_extract_all(alice, pattern)
```


