---
title: "Regular Expressions in R"
date: "`r lubridate::today()`"
output: 
  html_document:
    toc: true
---

## Definition and uses

Regular Expressions or "regexes" (but you will frequently hear also the singular name "regex"): **is a language for pattern matching** used by many other programming languages such as R, Python, etc.

Regular Expressions allow to manipulate texts. They **are strings containing normal characters and/or special meta-characters** that describe a particular pattern that we want to match in a given text.

Thus they are **especially handy for any task that deals with text**: NLP or data-cleaning tasks (e.g., find words that include a given set of letters, how often do past tenses occur in a text, find emails or phone numbers, find and replace HTML tags, etc.).

`stringr` and regular expressions will be particularly useful for those of you who plan to work with textual data in this course and beyond.  


## Regex with `stringr`: view functions

```{r}
library(stringr)
suppressPackageStartupMessages(library(tidyverse))
```

`stringr` is the most popular R package for regular expression. 

Use functions from this package in combination with regular expressions ([see the cheat sheet](https://evoldyn.gitlab.io/evomics-2018/ref-sheets/R_strings.pdf)) to extract text from strings. We begin with two helper functions from this package: 

* `str_view()` returns the first regex match
* `str_view_all()` returns all regex matches

These two functions take a string ((a character vector) and a regular expressions as inputs, and return the match. 

NB: these two functions allow only to VIEW the match, and we use them to show how regular expression works. If you need to replace or manipulate it, you need other functions from the `stringr`. The most common are:

* `str_count()`: count the number of matches
* `str_extract()` and `str_extract_all()`: extract matches
* `str_replace()` and `str_replace_all()`: replace matches
* `str_detect()`: detect matches in a string (return TRUE or FALSE)
* `str_split()`: split a string based on a match

These functions all work in a similar way: you pass the string and the regular expression as inputs, and they return a match, but they allow to do more than just viewing match (e.g., count, extract, replace them, etc.)


## Regex using normal characters

**Match specific characters or full words:**
```{r}
x <- c("This is a sentence to practice regular expressions", 
       "Find the word regex in this sentence",
       "A last sentence similar to the first sentence")
```

```{r}

```

This is fine, but regular expressions are more powerful than replacing manually specified characters or full words: they recognize **meta-characters** (see definition at the top), that is characters that have a unique meaning to the regex matching engine and vastly enhance the search capability.


## Most common regex meta-characters: `.` , `^`, `$`, `|`, `[]`

**The `.` matches any single character:**
```{r}

```

The `.` dot is a wildcard that matches any single character (except for newline characters); therefore, it is useful when you have a mix of characters (letters, digits, whitespaces, weird characters, etc.) or when you do not have detailed information about the text you are trying to match.

**The `^` caret anchors a match to the start of the string:**
```{r}

```

**The `$` dollar sign anchors a match to the end of the string:**
```{r}

```


**PRACTICE 1.** Use dots and `$` to match the last four characters of all sentences in `x`: 
```{r practice1-solution}
# your code here

```


**The `or` operator is represented using the `|` character:**
```{r}

```

The `or` operator can also be represented by grouping a set of characters using `[]`, which is shorthand for `|`. The main difference is that `[]` only works for single characters, not for entire words.

**Use `[]` to match any (one) of the characters in the `[]`** (e.g., [abc] matches a, b, or c):
```{r}

```

**Use `[]` with a `^` caret placed after the first `[` to match anything except what is in the `[]`** (e.g., [^abc] matches anything except a, b, or c):
```{r}

```

**If placed outside square brackets, the caret has a different meaning** that is checks whether one or more characters are located at the beginning of a strings:
```{r}

```

**The square brackets can be used to show any consecutive letters or numbers.** For example, use `[0-9]` to check whether a string contains any two consecutive decimal digit characters:

```{r}

```

```{r}

```

To check whether our string contains four alphabetic characters between the letter a and z, use `[a-z]` for lowercase and `[A-Z]` for uppercase characters:
```{r}

```


**PRACTICE 2.** This sentence contains a set of ugly characters. Identify them first using the `|` operator, then using the `[]`:
```{r}
j <- "This&is#MACS%30500"
```

```{r practice2-solution}
# your code for `|` here

# your code for [] here

```


## Additional regex meta-characters: `\d`, `\D`, `\w`, `\W`, `\s`, `\S`

Other useful meta-characters are `\d`, `\D`, `\w`, `\W`, `\s`, `\S`. Let's check how they work using the sentence below as example: 
```{r}
z <- "There are several assignments in this course: HW1, HW2, HW3, HW4, etc..."
```

**`\d` matches any decimal digit character, same as [0-9]**
```{r}

```

Note the `\\` double-backslashes: this is because to include a literal backslash (and the meta-character `\d` has one) you will need to double it. If you want to avoid them, pass the regular expression as a "raw string":
```{r}

```

**`\D` matches any character that is NOT decimal digit, same as [^0-9]**
```{r}

```

**`\w` matches any alphanumeric character, same as [a-zA-Z0-9]**
```{r}

```

**`\W` matches any character that is NOT alphanumeric, same as [^a-zA-Z0-9]**
```{r}

```

**`\s` matches any white space character**
```{r}

```

**`\S` matches any character that is NOT white space**
```{r}

```


**PRACTICE 3.** Apply these meta-characters to extract all phone numbers from the following string:

```{r}
s <- "My number is 444-3340, Carl number is 555-3755, Richard number is 666-6681"
```

```{r pratice3-solution}
# your code here
```


## Regex quantifiers: `*`, `+`, `?`, `{}` 

We can simply the above code using quantifiers. These are meta-characters that allow to **specify quantifiable conditions, such that a certain character must occur 0 or more times, 1 or more times, and so on**

The most common quantifiers are:

* `*` 	    matches 0 or more single character
* `+` 	    matches 1 or more
* `?` 	    matches 0 or 1
* `{2}`     matches a specified number of repetitions, here exactly 2
* `{2,5}`	  between 2 and 5 repetitions
* `{2,}`	  2 or more repetitions
* `{,5}`	  up to 5 repetitions

For example, we can use quantifiers to achieve the same results we did in practice 3 (extract all phone numbers from a string):
```{r}

```

Notice that quantifiers, like * or + or ?, apply to the character immediately to the left and not the full set of characters. 


**PRACTICE 4.** Match all three emails in the following string:
```{r}
email <- "smith90@uchicago.edu nardin@uchicago.edu happy2@uchicago.edu"
```

```{r pratice4-solution}
# your code here

```

**Quantifiers can be "greedy" or "lazy" ("non-greedy")**: by default, they are greedy.

Examples of Greedy quantifiers are `?`,`*`, `+`, and `{}`:

* They match as many characters as possible, and return the longest string at the first match that matches the pattern (e.g., the regex starts from the end of string, then backtracks until it matches the pattern)
* Example: the regex `a+` will match as many "a" as possible in the string "aaaa" even even though the substrings "a", "aa", "aaa" all match the regex "a+"

Examples of Lazy or Non-Greedy quantifiers are `??`, `*?`, `+?`, and `{}?`:

* They match as few characters as possible, and stop at the first recurrence of a character (e.g., the regex starts from the beginning of the string: moves forward through the string one character at a time, and stops at the first match)
* Example: the regex `a+?` will match as few "a" as possible in the string "aaaa". Thus, it matches the first character "a" and is done with it


## Regex with `stringr`: additional functions

We reviewed the most important regex meta-characters and quantifiers, and illustrated their use with examples and with `str_view()` and `str_view_all()`.
Below are additional examples to illustrate other common functions from `stringr`. Check the [documentation](https://stringr.tidyverse.org/) to review all arguments that these functions (and many more) can take.

Let's imagine we have a text, here the beginning of Alice in Wonderland, and we want to manipulate it. 
```{r}
alice <- "Alice was beginning to get very tired of sitting by her sister on the bank, and of having nothing to do: once or twice she had peeped into the book her sister was reading, but it had no pictures or conversations in it, 'and what is the use of a book,' thought Alice 'without pictures or conversation?'"
```

**`str_count()`: count the number of matches**
```{r}

```

**`str_extract()` and `str_extract_all()`: extract matches**
```{r}

```

**`str_replace()` and `str_replace_all()`: replace matches**
```{r}

```

```{r}

```

**`str_split()`: split a string based on a match**
```{r}

```

Matches are generally saved in a list. For example:
```{r}

```


## Resources

* [stringr cheat sheet](https://evoldyn.gitlab.io/evomics-2018/ref-sheets/R_strings.pdf) for a complete overview of all `stringr` functions
* [Chapter 14 "Strings" of R for Data Science](https://r4ds.had.co.nz/strings.html#strings), especially section 14.4 "Tools" for examples of each of these functions
* See slides and class websites (suggested readings) for more resources
